OPTIMAL PRICING AND INFORMAL SHARING: \\ EVIDENCE FROM PIPED WATER IN MANILA

### 2019-01-29 phil_paper/paper_wjv.tex
new motivation (pre-stef)

	In developing cities, high costs of connecting to piped water lead many households to use low-quality, non-piped sources.  In response, cities often set low fixed prices so that more households connect and high marginal prices to cover production costs.  This paper identifies a common alternative source --- sharing a neighbor's connection --- that has large implications for optimal water pricing.  With microdata from 1.5 million water connections in Manila, I structurally estimate water demand in the presence of water sharing and solve the regulator's problem: setting prices to maximize consumer surplus while funding production costs.  I find that low fixed and high marginal prices weaken incentives to share water between neighbors.  In contrast, the optimal pricing policy features high fixed and low marginal prices, increases shared connections, and ensures nearly universal access to piped water compared to current prices in Manila.

@@

\section{Introduction}
 % \footnote{\cite{laffont2005regulation} emphasizes high costs of public funds in developing countries (with a deadweight loss of 2.5 for the Philippines) due to limited ability to enter long-term contracts with government, credit constraints, inefficient taxation, and corruption.}
 % To ensure widespread access while covering costs, a 
Improving access to piped water can generate substantial health benefits\footnote{\cite{zhang2016long,kosec2014child,gamper2010impact}, and \cite{galiani2005water} focus on health improvements in developing countries while \cite{cutler2005role} look historically at the US.} as well as improvements in household well-being through time savings and reduced intra-household conflict.\footnote{\cite{devoto2012happiness} provide water through a randomized-controlled trial in Morocco finding few health effects but large benefits to household welfare.}  Yet, the share of urban households without piped water has remained stable at 5\% for middle income countries and at 20\% for low income countries (World Bank [2015]).  While external health and other benefits may merit subsidizing piped water, any subsidies are often limited by high costs of public funds and risk of regulatory capture in developing countries (\cite{laffont2005regulation}).  In many cases, prices must be set to fully recover costs.  To ensure access to piped water while covering costs, cities often subsidize upfront connection fees with high marginal prices for monthly usage.\footnote{\cite{mcintosh2003asian} and \cite{komives2006distributional} from the World Bank and Asian Development Bank argue for low connection fees.  Using a randomized controlled trial from Morocco, \cite{devoto2012happiness} confirm that modest assistance with connection fees can lead to increases in new water connections.}  Over 70\% of developing cities increase marginal prices with monthly usage so that large users cross-subsidize access and usage for smaller, poorer users.\footnote{This strategy is also commonly referred to as ``Increasing Block Tariffs'' or  ``Convex Pricing'' (\cite{borenstein2012redistributional}). \cite{hoque2013state} and \cite{boland2000political} provide overviews of the use of increasing marginal prices in the water sector in developing countries.  The statistic comes from author calculations with data kindly provided by \cite{hoque2013state}.}  These pricing approaches have been shown to work well when each water connection serves a single household.  Recent empirical papers in South Africa and C{\^o}te d'Ivoire predict gains in both access and total welfare from increasing marginal prices (\cite{szabo2015value,diakite2009proposal}).  However, many households in developing cities share connections with their neighbors, which can have large implications for water pricing.

% often increase the costs of purchasing a water connection particularly for disadvantaged households

Insecure land-tenure, credit-constraints, and administrative costs often make it difficult for disadvantaged households to purchase water connections (\cite{devoto2012happiness}).  Also, many cities maintain high upfront fees despite policy recommendations otherwise (\cite{hoque2013state}).  As a result, 70\% of African utilities report frequent water sharing between households (\cite{keener2010provision}) ranging from fetching from neighbors with hoses and buckets in Tangiers, Morocco (\cite{devoto2012happiness}) to drawing from single taps that each serve compounds in Kumasi, Ghana (\cite{whittington1992possible}).\footnote{\cite{nauges2006water} finds further evidence of widespread water sharing in Sri Lanka.}  While documenting the prevalence of these  sharing networks, research has not yet examined the importance of water sharing for optimal pricing policy.

>>>>

%	High costs of connecting to public utilities often

%	lead households to share a connection with their neighbors instead of purchasing their own connection.

	Households often share connections to public utilities with their neighbors in order to split set-up and maintenance costs.  This paper analyzes how sharing connections affects optimal pricing for water utilities.  With microdata from 1.5 million water connections in Manila, I estimate water demand in the presence of water sharing and solve for prices that maximize consumer surplus while funding production costs.  In practice, cities often set low fixed prices so that more households connect and high marginal prices to cover production costs.  Instead, the optimal policy uses high fixed prices to cover costs and low marginal prices to encourage greater consumption.  Sharing households split fixed prices and enjoy inexpensive piped water with their neighbors instead of using low-quality, non-piped water.  Counterfactual exercises find increases in shared connections, total piped water consumption, and social welfare from optimal prices compared to current prices in Manila.

%	Yet, I find that low fixed and high marginal prices also weaken incentives to share water between neighbors.

%	Shared connections may have large implications for optimal pricing for public utilities especially 
%	Households often offset the costs of connecting to public utilities by sharing a connection with their neighbors.  
%	Sharing neighbor's connections 
% In developing cities, high costs of connecting to piped water lead many households to use low-quality, non-piped sources.  
% In response, cities often set low fixed prices so that more households connect and high marginal prices to cover production costs.  
% This paper identifies a common alternative source --- sharing a neighbor's connection --- that has large implications for optimal water pricing.

@@

\section{Introduction}
 % \footnote{\cite{laffont2005regulation} emphasizes high costs of public funds in developing countries (with a deadweight loss of 2.5 for the Philippines) due to limited ability to enter long-term contracts with government, credit constraints, inefficient taxation, and corruption.}
 % To ensure widespread access while covering costs, a

%  For example, when households can easily share connections, public utilities can charge larger upfront fees for the same levels of access as when sharing is .  [UNCLEAR COMPARISON!!]

% Households also share connections to public utilities with their neighbors (piped water, electricity, mobile phones, etc.) especially in developing countries.\footnote{70\% of African water utilities report frequent sharing of water taps between households.  Over 14\% of urban households in Arba-Minch, Ethiopia report sourcing electricity from their neighbors instead of purchasing directly from the utility (\cite{mustefa2016energy}).  In South Africa as of 2012, 11.9\% of the population report sharing their mobile phone with others.  This percentage jumps to 19.9\% among those living in poverty (\cite{calandro2012mobile}).}

% Connection sharing can have important implications for the optimal balance of fixed and marginal prices for these services.

% Understanding how this sharing behavior may affect optimal prices may have large consequences for public policy since governments actively regulate prices charged by public utilities.

% In developing countries, households often 
% Faced with fixed costs of accessing a good or service, consumers are increasingly .  
% In developing countries, these .

Faced with high fixed costs of purchasing and maintaining durable goods, consumers often share these goods with each other.  While online platforms have accelerated peer-to-peer rentals of cars, apartments, and storage space (\cite{horton2016owning}), households in developing countries often share connections to public utilities like water and electricity with their neighbors.\footnote{70\% of African water utilities report frequent sharing of water taps between households.  Over 14\% of urban households in Arba-Minch, Ethiopia report sourcing electricity from their neighbors instead of purchasing directly from the utility (\cite{mustefa2016energy}).  In South Africa as of 2012, 11.9\% of the population report sharing their mobile phone with others.  This percentage jumps to 19.9\% among those living in poverty (\cite{calandro2012mobile}).}  Sharing can have large consequences for optimal pricing, particularly for public utilities.  Public utility pricing is especially salient in developing countries because prices must often be set to fully cover costs (\cite{laffont2005regulation}) and because prices determine not only monthly consumption, but also which households can access these basic services.  For example, when households cannot share connections, they may prefer relatively low fixed prices (ie. installation and maintenance fees) so that each household can afford a connection and high marginal prices (ie. prices per gallon or watt) to cover production costs.  However, when households can easily share connections, they may prefer splitting high fixed prices with their neighbors and enjoying low marginal prices from shared connections.  
%The optimal pricing policy depends both on consumer demand and on the costs of sharing connections between neighbors.

This paper focuses empirically on pricing piped water in Manila, Philippines to increase access and welfare when households share connections.  Improving access to piped water can generate substantial health benefits\footnote{\cite{zhang2016long,kosec2014child,gamper2010impact}, and \cite{galiani2005water} focus on health improvements in developing countries while \cite{cutler2005role} look historically at the US.} as well as improvements in household well-being through time savings and reduced intra-household conflict.\footnote{\cite{devoto2012happiness} provide water through a randomized-controlled trial in Morocco finding few health effects but large benefits to household welfare.}  Yet, the share of urban households without piped water has remained stable at 5\% for middle income countries and at 20\% for low income countries (World Bank [2015]).  While external health and other benefits may merit subsidizing piped water, any subsidies to public utilities are often limited by high costs of public funds and risk of regulatory capture in developing countries (\cite{laffont2005regulation}).

To ensure access to piped water while covering costs, cities often subsidize upfront connection fees with high marginal prices for monthly usage.\footnote{\cite{mcintosh2003asian} and \cite{komives2006distributional} from the World Bank and Asian Development Bank argue for low connection fees.  Using a randomized controlled trial from Morocco, \cite{devoto2012happiness} confirm that modest assistance with connection fees can lead to increases in new water connections.}  Over 70\% of developing cities increase marginal prices with monthly usage so that large users cross-subsidize access and usage for smaller, poorer users.\footnote{This strategy is also commonly referred to as ``Increasing Block Tariffs'' or  ``Convex Pricing'' (\cite{borenstein2012redistributional}). \cite{hoque2013state} and \cite{boland2000political} provide overviews of the use of increasing marginal prices in the water sector in developing countries.  The statistic comes from author calculations with data kindly provided by \cite{hoque2013state}.}  These pricing approaches have been shown to work well when each water connection serves a single household.  Recent empirical papers in C{\^o}te d'Ivoire and South Africa predict gains in both access and total welfare from increasing marginal prices (\cite{diakite2009proposal,szabo2015value}).  However, many households in developing cities share connections with their neighbors, which can have large implications for water pricing.

Insecure land-tenure, credit-constraints, and administrative costs often make it difficult for disadvantaged households to purchase their own water connections (\cite{devoto2012happiness}).  As a result, households frequently share connections ranging from fetching from neighbors with hoses and buckets in Tangiers, Morocco (\cite{devoto2012happiness}) to drawing from single taps that each serve compounds in Kumasi, Ghana (\cite{whittington1992possible}).\footnote{\cite{nauges2006water} finds further evidence of widespread water sharing in Sri Lanka.}  While documenting the prevalence of these  sharing networks, research has not yet examined the importance of water sharing for optimal pricing policy.

% often increase the costs of purchasing a water connection particularly for disadvantaged households
% Also, many cities maintain high upfront fees despite policy recommendations otherwise (\cite{hoque2013state}).
